\dppTitle

### 2014-03-07 manuscripts/dpp-msbayes/dpp-msbayes-main.tex
Updating references in dpp ms.
Editing sentences where `citet` was previously used. Making these
sentences compatible with the current use of `cite`.

Also, adding HKY85 reference, and updating natbib to sort and compress
numerical citations.

Recently, \cite{Oaks2012} and \cite{Oaks2014reply} found via simulation that
\msb will often strongly support biogeographically interesting models with a
small number of divergence events shared among taxa, even when divergences were
random over broad timescales.
They suggested this behavior was due to the combination of uniform priors on
parameters causing small marginal likelihoods of richer models and the
U-shaped prior on divergence models.
\cite{Hickerson2013} suggested the problem is caused by sampling error, and
proposed as a solution an approximate-Bayesian model averaging approach that
samples over empirically informed uniform priors.
However, \cite{Oaks2014reply} evaluated the approach proposed by
\cite{Hickerson2013} using simulations and found that it did not mitigate the
method's propensity to incorrectly infer clustered divergences, and often
preferred priors that excluded the true values of the model's parameters.
Here, I describe a new method that successfully mitigates spurious inference
of biogeographically interesting scenarios while avoiding negative side effects
of empirically informed uniform priors.

@@

\subsection{The model}
In this section, I describe the model, which is a modification of the model
implemented in \msb \cite{Huang2011,Oaks2012}.
The code implementing the new model is freely available in the open-source
software package \dppmsbayes
(\href{https://github.com/joaks1/dpp-msbayes}{\url{https://github.com/joaks1/dpp-msbayes}}).
To perform the analyses described below, I used the freely avaliable,
open-source software package \pymsbayes
(\href{https://github.com/joaks1/PyMsBayes}{\url{https://github.com/joaks1/PyMsBayes}}),
which is a Python API that provides a multi-processing interface to \msb and
\dppmsbayes.
I performed the work described below following the principles of
Open Notebook Science.
Using version-control software, I make progress in all aspects of the work
freely and publicly available in real-time at
\href{https://github.com/joaks1/msbayes-experiments}{\url{https://github.com/joaks1/msbayes-experiments}}.
All information necessary to reproduce my results is provided there.
I follow much of the notation of \cite{Oaks2012}, but modify it to
aid in the description of the new model.
A summary of my notation can be found in Table \ref{tabNotation}.

@@

\subsubsection{Likelihood terms of Equation \ref{eq:fullModelCompact}}
\begin{linenomath}
The likelihood terms of Equation \ref{eq:fullModelCompact} can
be expanded out as a product over population pairs and loci
\begin{equation}
    p(\alignmentVector \given \geneTreeVector, \hkyModelVector)
    p(\geneTreeVector \given \divTimeMapVector, \demographicParamVector,
    \locusMutationRateScalarVector, \ploidyScalarVector,
    \mutationRateScalarConstantVector) = 
    \prod_{i=1}^{\npairs{}}
    \prod_{j=1}^{\nloci{i}}
    p(\alignment{i}{j} \given \geneTree{i}{j}, \hkyModel{i}{j})
    p(\geneTree{i}{j} \given \divTimeMap{i}, \demographicParams{i},
    \locusMutationRateScalar{j}, \ploidyScalar{i}{j},
    \mutationRateScalarConstant{i}{j}).
    \label{eq:modelLikelihoodExpanded}
\end{equation}
The first term,
$p(\alignment{i}{j} \given \geneTree{i}{j}, \hkyModel{i}{j})$,
is the probability of the sequence alignment of locus $j$ for population pair
$i$ given the gene tree and \hky substitution model parameters \cite[i.e., the
``Felsenstein likelihood'';]{Felsenstein1981}.
The model allows for an intra-locus recombination rate \recombinationRate,
which, for simplicity, is assumed to be zero in Equation
\ref{eq:fullModelCompact}.
If \recombinationRate is non-zero, this term requires an additional product
over the columns (sites) of each sequence alignment to allow sites to have
different genealogies.
The second term,
p(\geneTree{i}{j} \given \divTimeMap{i}, \demographicParams{i},
\locusMutationRateScalar{j}, \ploidyScalar{i}{j},
\mutationRateScalarConstant{i}{j}),
is the probability of the gene tree under a multi-population coalescent model
(i.e., species tree) where the ancestral population of pair $i$ diverges
and gives rise to the two sampled descendant populations.
Each \demographicParams{} contains the following demographic parameters: The
mutation-rate-scaled effective sizes ($\myTheta{} = 4N\mutationRate$) of the
ancestral, \ancestralTheta{}, and descendant populations, \descendantTheta{1}{}
and \descendantTheta{2}{};
the proportion of the first, \bottleScalar{1}{}, and second population,
\bottleScalar{2}{}, that persist during bottlenecks that begin immediately
after divergence in forward-time;
the proportion of time between present and divergence when the bottlenecks
end for both populations, \bottleTime{};
and the symmetric migration rate between the descendant populations,
\migrationRate{}.
Thus, the probability of the $\popSampleSize{i}{}-1$ coalescence times (node
heights) of gene tree \geneTree{i}{j} is given by a multi-population
Kingman-coalescent model \cite{Kingman1982} where the ancestral population of
size
$\ancestralTheta{i}\ploidyScalar{i}{j}\mutationRateScalarConstant{i}{j}\locusMutationRateScalar{j}$
diverges at time \divTimeMap{i} into two descendant populations of constant
size
$\descendantTheta{1}{i}\ploidyScalar{i}{j}\mutationRateScalarConstant{i}{j}\locusMutationRateScalar{j}\bottleScalar{1}{i}$
and
$\descendantTheta{2}{i}\ploidyScalar{i}{j}\mutationRateScalarConstant{i}{j}\locusMutationRateScalar{j}\bottleScalar{2}{i}$,
which, after time $\divTimeMap{i}\bottleTime{i}$, grow exponentially to their
present size 
$\descendantTheta{1}{i}\ploidyScalar{i}{j}\mutationRateScalarConstant{i}{j}\locusMutationRateScalar{j}$
and
$\descendantTheta{2}{i}\ploidyScalar{i}{j}\mutationRateScalarConstant{i}{j}\locusMutationRateScalar{j}$,
respectively.
Following divergence, the descendant populations of pair $i$ exchange migrants at a symmetric rate of
\migrationRate{i}.
\end{linenomath}

@@

\begin{linenomath}
As described by \cite{Oaks2012}, to get the divergence times in units
proportional to the expected number of mutations, they must be
scaled by the realized population size for locus $j$ of population-pair $i$ 
\begin{equation}
    \divTimeScaled{i}{j} = \divTimeMap{i} \times \frac{\myTheta{C}}{
        \descendantThetaMean{i} \ploidyScalar{i}{j}},
    \label{eq:divTimeScaled}
\end{equation}
where \descendantThetaMean{i} is the mean of \descendantTheta{1}{} and
\descendantTheta{2}{} for pair $i$.
This gives us the vector of scaled divergence times
$\divTimeScaledVector = (\divTimeScaled{1}{1}, \ldots,
\divTimeScaled{\npairs{}}{\nloci{\npairs{}}})$.
\end{linenomath}

@@

\subsection{Differences between this model and the original \texttt{msBayes} model}
\subsubsection{The prior on divergence models}
One of the key differences between my model and that of \msb \cite{Huang2011}
is the prior distribution on divergence models.
As discussed in \cite{Oaks2012}, in \msb the prior used for
\divTimeIndexVector is a combination of a discrete uniform prior over the
possible number of divergence events \divTimeNum from 1 to \npairs{} with a
multinomial distribution on the number of times each index of \divTimeVector
appears in \divTimeIndexVector, with the constraint that all \divTime{}
parameters are represented at least once (see Equation 2 of \cite{Oaks2012}).
I denote this prior used by \msb as $\divTimeIndexVector \sim \priorOld$.
\cite{Oaks2012} discuss how placing a uniform prior over the number of
divergence parameters (denoted \divTimeNum here, and as \numt{} in
\cite{Huang2011}) imposes an U-shaped prior over divergence
models (\divTimeIndexVector; see
Figure 5B of \cite{Oaks2012}).
To avoid this, I place priors directly on the sample space of divergence
models, thus eliminating the parameter \numt{} from the model.
I introduce two priors on divergence models:
(1) a prior that is uniform over all unordered divergence models, and
(2) a Dirichlet-process prior on all ordered divergence models.
The latter provides an investigator with a great deal of flexibility in
expressing their prior beliefs about models of divergence.

@@

In addition, I have modified the prior on the sizes of the descendant
populations of each pair.
As described by \cite{Oaks2012}, \msb uses the joint prior
\begin{equation}
    \descendantTheta{1}{}, \descendantTheta{2}{} \sim
    Beta(1,1) \times 2 \times U(\uniformMin{\myTheta{}},
    \uniformMax{\descendantTheta{}{}}),
    \label{eq:jointThetaPrior}
\end{equation}
such that the user-specified uniform prior on descendant population
size is a prior on the \emph{mean} size of the two descendant
populations of each pair.
Under my model, the sizes of the descendant populations of each
pair are \iid as
$\descendantTheta{1}{} \sim Gamma(\gammaShape{\descendantTheta{}{}},
\gammaScale{\descendantTheta{}{}})$
and
$\descendantTheta{2}{} \sim Gamma(\gammaShape{\descendantTheta{}{}},
\gammaScale{\descendantTheta{}{}})$.
This relaxes the assumption that the sizes of the two descendant populations
are interdependent and negatively correlated.

@@

\subsubsection{Time scale}
As described above, divergence times are in units of
$\myTheta{C}/\mutationRate$ generations, where \myTheta{C} is the expectation
of the prior on descendant-population size.
As described by \cite{Oaks2012}, in \msb, \myTheta{C} is half of the upper
limit of the continuous uniform prior on the mean of the descendant population
sizes.
This is only equal to the expectation of the prior if the lower limit is zero.

@@

\subsubsection{Ordering of taxon-specific summary statistics}
As alluded to in the model description, \msb does not maintain the order of the
taxon-specific summary statistics \alignmentSS{}{} within each \ssVector{}.
Rather, the summary statistics are re-ordered by descending values of average
pairwise differences between the descendant populations
\cite[$\pi_b$;]{NeiLi1979,Huang2011}.
This has the advantage of reducing the sample space of the number of
divergence models \divTimeIndexVector, but there are at least two disadvantages.
First, additional information in the data is lost.
By discarding the identity of the \npairs{} pairs of populations, all
pair-specific information about the amount of data (e.g., the number gene
copies collected from each of the populations [\popSampleSize{1}{} and
\popSampleSize{2}{}], the number of loci, and the length of the loci), and the
taxon- and locus-specific parameters (\hkyModel{}{},
\mutationRateScalarConstant{}{}, \ploidyScalar{}{}, and
\locusMutationRateScalar{}) is lost.
Second, the results are more difficult to interpret, because they can be no
longer be directly associated to the taxa under study.

@@

\subsubsection{Obtaining an approximate posterior from the prior samples}
I use a rejection algorithm to retain an approximate posterior sample of
\paramSampleVector{} from the prior sample
$\paramSampleMatrix = (\paramSampleVector{1}, \ldots, \paramSampleVector{\numPriorSamples})$.
First, the observed summary statistics \ssVectorObs, and the summary statistics
of the prior samples 
$\ssMatrix = (\ssVector{1}, \ldots, \ssVector{\numPriorSamples})$,
are standardized using the means and standard deviations of the statistics from
the prior sample (i.e., the prior mean is subtracted from each statistic, and the
difference is divided by the prior standard deviation).
After all statistics are standardized, the Euclidean distance between
\ssVectorObs and each \ssVector{} within \ssMatrix is calculated.
The samples that fall within a range of tolerance \tol around \ssVectorObs
are retained.
The range of tolerance is determined by specifying the desired number of
posterior samples to be retained.
Post-hoc adjustment of the posterior sample can also be performed with a number
of regression techniques \cite{Beaumont2002,Blum2009,Leuenberger2010}.
For my analyses, I use the general linear model (GLM) regression adjustment
\cite{Leuenberger2010} as implemented in \abctoolbox
\cite[v1.1;]{ABCtoolbox}, which \cite{Oaks2012} showed performs very
similarly to weighted local-linear regression and multinomial logistic
regression adjustments \cite{Beaumont2002} for \msb posteriors.

\subsection{Assessing model-choice behavior and robustness}
Following the simulation-based approach of \cite{Oaks2012}, I characterize
the behavior of several models under the ideal conditions where the
data are generated from parameters drawn from the same prior distributions used
for analysis (i.e., the prior is correct).
I selected the following four model priors for these analyses
(Table~\ref{tabPriors}).
\begin{enumerate}
    \item The \modelOld model represents the original \msb implementation with
        the U-shaped prior on unordered divergence models and uniform priors on
        divergence-time and demographic parameters; $\divTimeIndexVector \sim
        \priorOld$, $\divTime{} \sim U(0,10)$, $\ancestralTheta{} \sim U(0,
        0.05)$, and $\descendantThetaMean{} \sim U(0, 0.05)$.
    \item The \modelUshaped model with the U-shaped prior of \msb on unordered
        divergence models, but with exponential priors on divergence-time and
        demographic parameters; $\divTimeIndexVector \sim \priorOld$,
        $\divTime{} \sim Exp(mean=2.887)$, $\ancestralTheta{} \sim
        Exp(mean=0.025)$, $\descendantTheta{1}{} \sim Exp(mean=0.025)$, and
        $\descendantTheta{2}{} \sim Exp(mean=0.025)$.
    \item The \modelUniform model with a uniform prior over unordered
        divergence models and exponential priors on divergence-time and
        demographic parameters; $\divTimeIndexVector \sim \priorUniform$,
        $\divTime{} \sim Exp(mean=2.887)$, $\ancestralTheta{} \sim
        Exp(mean=0.025)$, $\descendantTheta{1}{} \sim Exp(mean=0.025)$, and
        $\descendantTheta{2}{} \sim Exp(mean=0.025)$.
    \item The \modelDPP model with a Dirichlet-process prior on ordered
        divergence models and exponential priors on divergence-time and
        demographic parameters; $\divTimeIndexVector \sim \priorDPP{\sim
        Gamma(2,2)}$, $\divTime{} \sim Exp(mean=2.887)$, $\ancestralTheta{}
        \sim Exp(mean=0.025)$, $\descendantTheta{1}{} \sim Exp(mean=0.025)$,
        and $\descendantTheta{2}{} \sim Exp(mean=0.025)$.
\end{enumerate}
I selected the exponential prior on divergence time used in models \modelDPP,
\modelUniform, and \modelUshaped to have the same variance as the uniform prior
in model \modelOld.
I selected the exponential prior on population size used in models \modelDPP,
\modelUniform, and \modelUniform to have the same mean as the uniform prior in
model \modelOld, so that all four models have the same \myTheta{C} and thus the
same units of time.
All of the models were the same in other respects, with three free \myTheta{}
parameters for each population pair, two uniformly distributed ($beta(1,1)$)
\bottleScalar{}{} parameters per pair, no migration, no recombination, and
resorting of taxon-specific summary statistics by $\pi_b$ (i.e., sampling
unordered divergence models).
For all simulations, I used a data structure of eight population pairs, with a
single 1000 base-pair locus sampled from 10 individuals from each population.

@@

For each set of 50,000 simulated datasets, I used the posterior estimates
to assess the model-choice behavior of each model.
I did this by assigning the 50,000 estimates of the posterior probability
of one-divergence event to 20 bins of width 0.05, and plotted
the estimated posterior probability of each bin against the proportion of
replicates in that bin with a true value consistent with one divergence
event \cite{Huelsenbeck2004,Oaks2012}.
If the method is unbiased, the points should fall near the identity line,
whereas any trend away from the identity line indicates bias.
I did this using two criteria for the one-divergence model:
(1) the number of divergence-time parameters ($\divTimeNum = 1$) and
(2) the dispersion index of divergence times ($\divTimeDispersion < 0.01$).
For the latter, $\divTimeDispersion < 0.01$ has been commonly used as an
arbitrary criterion for a single ``simultaneous'' divergence event
\cite[e.g.,]{Hickerson2006,Leache2007,Stone2012}.
I focused on the one-divergence model to assess model-choice behavior, because
it is often of biogeographic interest and is easily comparable among
the three different priors used on divergence models.

@@

\subsection{Assessing power}
I evaluated the power of the same four models (Table~\ref{tabPriors}) to
detect random variation in divergence times using methods similar to
\cite{Oaks2012}.
For all power simulations, I used a data structure identical to that of the
empirical dataset of Philippine vertebrates analyzed by \cite{Oaks2012}, which
consists of 22 pairs of populations.
Due to the larger number of pairs, I used a different hyperprior on the
concentration parameter for the \modelDPP model; I used a prior of
$\divTimeIndexVector \sim \priorDPP{\sim Gamma(1.5,18.1)}$ over divergence
models for the model \modelDPP.
All other aspects of the four models in Table~\ref{tabPriors} were identical to
those used in the validation analyses described above.
For each of the four models, I generated $2\e6$ samples from the prior.

@@

\subsection{Validation analyses: Model-choice accuracy}
The \msb model, and my modification of it, is a model-choice method
with the primary purpose of estimating the probabilities of models
of divergence across taxa.
Thus, it is critical to assess the method's ability to accurately
estimate the posterior probabilities of divergence models.
Consistent with the findings of \cite{Oaks2012}, my results demonstrate that
the unadjusted estimates of divergence-model posterior probabilities are
generally more accurate than regression-adjusted estimates
(compare the plots along the upper-left to lower-right diagonal for
Figure~\ref{figValidationModelChoicePsi} versus
S\ref{figValidationModelChoicePsiGlm} and
Figure~\ref{figValidationModelChoiceOmega} versus
S\ref{figValidationModelChoiceOmegaGlm}).
Regression adjustment results in biased estimates of the posterior probability
of the one-divergence model when all model assumptions are satisfied, which is
well illustrated in Figure S\ref{figValidationModelChoiceOmegaGlm}.
As a result, I will focus my discussion of the results on the unadjusted
estimates.

>>>>

Recently, Oaks et al.\ \cite{Oaks2012,Oaks2014reply} found via simulation that
\msb will often strongly support biogeographically interesting models with a
small number of divergence events shared among taxa, even when divergences were
random over broad timescales.
They suggested this behavior was due to the combination of uniform priors on
parameters causing small marginal likelihoods of richer models and the
U-shaped prior on divergence models.
Hickerson et al.\ \cite{Hickerson2013} suggested the problem is caused by
sampling error, and proposed as a solution an approximate-Bayesian model
averaging approach that samples over empirically informed uniform priors.
However, Oaks et al.\ \cite{Oaks2014reply} evaluated the approach proposed by
Hickerson et al.\ \cite{Hickerson2013} using simulations and found that it did
not mitigate the method's propensity to incorrectly infer clustered
divergences, and often preferred priors that excluded the true values of the
model's parameters.
Here, I describe a new method that successfully mitigates spurious inference
of biogeographically interesting scenarios while avoiding negative side effects
of empirically informed uniform priors.

@@

\subsection{The model}
In this section, I describe the model, which is a modification of the model
implemented in \msb \cite{Huang2011,Oaks2012}.
The code implementing the new model is freely available in the open-source
software package \dppmsbayes
(\href{https://github.com/joaks1/dpp-msbayes}{\url{https://github.com/joaks1/dpp-msbayes}}).
To perform the analyses described below, I used the freely avaliable,
open-source software package \pymsbayes
(\href{https://github.com/joaks1/PyMsBayes}{\url{https://github.com/joaks1/PyMsBayes}}),
which is a Python API that provides a multi-processing interface to \msb and
\dppmsbayes.
I performed the work described below following the principles of
Open Notebook Science.
Using version-control software, I make progress in all aspects of the work
freely and publicly available in real-time at
\href{https://github.com/joaks1/msbayes-experiments}{\url{https://github.com/joaks1/msbayes-experiments}}.
All information necessary to reproduce my results is provided there.
I follow much of the notation of Oaks et al.\ \cite{Oaks2012}, but modify it to
aid in the description of the new model.
A summary of my notation can be found in Table \ref{tabNotation}.

@@

\subsubsection{Likelihood terms of Equation \ref{eq:fullModelCompact}}
\begin{linenomath}
The likelihood terms of Equation \ref{eq:fullModelCompact} can
be expanded out as a product over population pairs and loci
\begin{equation}
    p(\alignmentVector \given \geneTreeVector, \hkyModelVector)
    p(\geneTreeVector \given \divTimeMapVector, \demographicParamVector,
    \locusMutationRateScalarVector, \ploidyScalarVector,
    \mutationRateScalarConstantVector) = 
    \prod_{i=1}^{\npairs{}}
    \prod_{j=1}^{\nloci{i}}
    p(\alignment{i}{j} \given \geneTree{i}{j}, \hkyModel{i}{j})
    p(\geneTree{i}{j} \given \divTimeMap{i}, \demographicParams{i},
    \locusMutationRateScalar{j}, \ploidyScalar{i}{j},
    \mutationRateScalarConstant{i}{j}).
    \label{eq:modelLikelihoodExpanded}
\end{equation}
The first term,
$p(\alignment{i}{j} \given \geneTree{i}{j}, \hkyModel{i}{j})$, is the
probability of the sequence alignment of locus $j$ for population pair $i$
given the gene tree and \hky \cite{HKY} substitution model parameters
\cite[i.e., the ``Felsenstein likelihood'']{Felsenstein1981}.
The model allows for an intra-locus recombination rate \recombinationRate,
which, for simplicity, is assumed to be zero in Equation
\ref{eq:fullModelCompact}.
If \recombinationRate is non-zero, this term requires an additional product
over the columns (sites) of each sequence alignment to allow sites to have
different genealogies.
The second term,
p(\geneTree{i}{j} \given \divTimeMap{i}, \demographicParams{i},
\locusMutationRateScalar{j}, \ploidyScalar{i}{j},
\mutationRateScalarConstant{i}{j}),
is the probability of the gene tree under a multi-population coalescent model
(i.e., species tree) where the ancestral population of pair $i$ diverges
and gives rise to the two sampled descendant populations.
Each \demographicParams{} contains the following demographic parameters: The
mutation-rate-scaled effective sizes ($\myTheta{} = 4N\mutationRate$) of the
ancestral, \ancestralTheta{}, and descendant populations, \descendantTheta{1}{}
and \descendantTheta{2}{};
the proportion of the first, \bottleScalar{1}{}, and second population,
\bottleScalar{2}{}, that persist during bottlenecks that begin immediately
after divergence in forward-time;
the proportion of time between present and divergence when the bottlenecks
end for both populations, \bottleTime{};
and the symmetric migration rate between the descendant populations,
\migrationRate{}.
Thus, the probability of the $\popSampleSize{i}{}-1$ coalescence times (node
heights) of gene tree \geneTree{i}{j} is given by a multi-population
Kingman-coalescent model \cite{Kingman1982} where the ancestral population of
size
$\ancestralTheta{i}\ploidyScalar{i}{j}\mutationRateScalarConstant{i}{j}\locusMutationRateScalar{j}$
diverges at time \divTimeMap{i} into two descendant populations of constant
size
$\descendantTheta{1}{i}\ploidyScalar{i}{j}\mutationRateScalarConstant{i}{j}\locusMutationRateScalar{j}\bottleScalar{1}{i}$
and
$\descendantTheta{2}{i}\ploidyScalar{i}{j}\mutationRateScalarConstant{i}{j}\locusMutationRateScalar{j}\bottleScalar{2}{i}$,
which, after time $\divTimeMap{i}\bottleTime{i}$, grow exponentially to their
present size 
$\descendantTheta{1}{i}\ploidyScalar{i}{j}\mutationRateScalarConstant{i}{j}\locusMutationRateScalar{j}$
and
$\descendantTheta{2}{i}\ploidyScalar{i}{j}\mutationRateScalarConstant{i}{j}\locusMutationRateScalar{j}$,
respectively.
Following divergence, the descendant populations of pair $i$ exchange migrants at a symmetric rate of
\migrationRate{i}.
\end{linenomath}

@@

\begin{linenomath}
As described by Oaks et al.\ \cite{Oaks2012}, to get the divergence times in
units proportional to the expected number of mutations, they must be scaled by
the realized population size for locus $j$ of population-pair $i$ 
\begin{equation}
    \divTimeScaled{i}{j} = \divTimeMap{i} \times \frac{\myTheta{C}}{
        \descendantThetaMean{i} \ploidyScalar{i}{j}},
    \label{eq:divTimeScaled}
\end{equation}
where \descendantThetaMean{i} is the mean of \descendantTheta{1}{} and
\descendantTheta{2}{} for pair $i$.
This gives us the vector of scaled divergence times
$\divTimeScaledVector = (\divTimeScaled{1}{1}, \ldots,
\divTimeScaled{\npairs{}}{\nloci{\npairs{}}})$.
\end{linenomath}

@@

\subsection{Differences between this model and the original \texttt{msBayes} model}
\subsubsection{The prior on divergence models}
One of the key differences between my model and that of \msb \cite{Huang2011}
is the prior distribution on divergence models.
As discussed in Oaks et al.\ \cite{Oaks2012}, in \msb the prior used for
\divTimeIndexVector is a combination of a discrete uniform prior over the
possible number of divergence events \divTimeNum from 1 to \npairs{} with a
multinomial distribution on the number of times each index of \divTimeVector
appears in \divTimeIndexVector, with the constraint that all \divTime{}
parameters are represented at least once (see Equation 2 of \cite{Oaks2012}).
I denote this prior used by \msb as $\divTimeIndexVector \sim \priorOld$.
Oaks et al.\ \cite{Oaks2012} discuss how placing a uniform prior over the
number of divergence parameters (denoted \divTimeNum here, and as \numt{} in
\cite{Huang2011}) imposes an U-shaped prior over divergence
models (\divTimeIndexVector; see
Figure 5B of \cite{Oaks2012}).
To avoid this, I place priors directly on the sample space of divergence
models, thus eliminating the parameter \numt{} from the model.
I introduce two priors on divergence models:
(1) a prior that is uniform over all unordered divergence models, and
(2) a Dirichlet-process prior on all ordered divergence models.
The latter provides an investigator with a great deal of flexibility in
expressing their prior beliefs about models of divergence.

@@

In addition, I have modified the prior on the sizes of the descendant
populations of each pair.
As described by Oaks et al.\ \cite{Oaks2012}, \msb uses the joint prior
\begin{equation}
    \descendantTheta{1}{}, \descendantTheta{2}{} \sim
    Beta(1,1) \times 2 \times U(\uniformMin{\myTheta{}},
    \uniformMax{\descendantTheta{}{}}),
    \label{eq:jointThetaPrior}
\end{equation}
such that the user-specified uniform prior on descendant population
size is a prior on the \emph{mean} size of the two descendant
populations of each pair.
Under my model, the sizes of the descendant populations of each
pair are \iid as
$\descendantTheta{1}{} \sim Gamma(\gammaShape{\descendantTheta{}{}},
\gammaScale{\descendantTheta{}{}})$
and
$\descendantTheta{2}{} \sim Gamma(\gammaShape{\descendantTheta{}{}},
\gammaScale{\descendantTheta{}{}})$.
This relaxes the assumption that the sizes of the two descendant populations
are interdependent and negatively correlated.

@@

\subsubsection{Time scale}
As described above, divergence times are in units of
$\myTheta{C}/\mutationRate$ generations, where \myTheta{C} is the expectation
of the prior on descendant-population size.
As described by Oaks et al.\ \cite{Oaks2012}, in \msb, \myTheta{C} is half of
the upper limit of the continuous uniform prior on the mean of the descendant
population sizes.
This is only equal to the expectation of the prior if the lower limit is zero.

@@

\subsubsection{Ordering of taxon-specific summary statistics}
As alluded to in the model description, \msb does not maintain the order of the
taxon-specific summary statistics \alignmentSS{}{} within each \ssVector{}.
Rather, the summary statistics are re-ordered by descending values of average
pairwise differences between the descendant populations
($\pi_b$) \cite{NeiLi1979,Huang2011}.
This has the advantage of reducing the sample space of the number of
divergence models \divTimeIndexVector, but there are at least two disadvantages.
First, additional information in the data is lost.
By discarding the identity of the \npairs{} pairs of populations, all
pair-specific information about the amount of data (e.g., the number gene
copies collected from each of the populations [\popSampleSize{1}{} and
\popSampleSize{2}{}], the number of loci, and the length of the loci), and the
taxon- and locus-specific parameters (\hkyModel{}{},
\mutationRateScalarConstant{}{}, \ploidyScalar{}{}, and
\locusMutationRateScalar{}) is lost.
Second, the results are more difficult to interpret, because they can be no
longer be directly associated to the taxa under study.

@@

\subsubsection{Obtaining an approximate posterior from the prior samples}
I use a rejection algorithm to retain an approximate posterior sample of
\paramSampleVector{} from the prior sample
$\paramSampleMatrix = (\paramSampleVector{1}, \ldots, \paramSampleVector{\numPriorSamples})$.
First, the observed summary statistics \ssVectorObs, and the summary statistics
of the prior samples 
$\ssMatrix = (\ssVector{1}, \ldots, \ssVector{\numPriorSamples})$,
are standardized using the means and standard deviations of the statistics from
the prior sample (i.e., the prior mean is subtracted from each statistic, and the
difference is divided by the prior standard deviation).
After all statistics are standardized, the Euclidean distance between
\ssVectorObs and each \ssVector{} within \ssMatrix is calculated.
The samples that fall within a range of tolerance \tol around \ssVectorObs
are retained.
The range of tolerance is determined by specifying the desired number of
posterior samples to be retained.
Post-hoc adjustment of the posterior sample can also be performed with a number
of regression techniques \cite{Beaumont2002,Blum2009,Leuenberger2010}.
For my analyses, I use the general linear model (GLM) regression adjustment
\cite{Leuenberger2010} as implemented in \abctoolbox
\cite[v1.1;]{ABCtoolbox}, which Oaks et al.\ \cite{Oaks2012} showed performs
very similarly to weighted local-linear regression and multinomial logistic
regression adjustments \cite{Beaumont2002} for \msb posteriors.

\subsection{Assessing model-choice behavior and robustness}
Following the simulation-based approach of Oaks et al.\ \cite{Oaks2012}, I
characterize the behavior of several models under the ideal conditions where
the data are generated from parameters drawn from the same prior distributions
used for analysis (i.e., the prior is correct).
I selected the following four model priors for these analyses
(Table~\ref{tabPriors}).
\begin{enumerate}
    \item The \modelOld model represents the original \msb implementation with
        the U-shaped prior on unordered divergence models and uniform priors on
        divergence-time and demographic parameters; $\divTimeIndexVector \sim
        \priorOld$, $\divTime{} \sim U(0,10)$, $\ancestralTheta{} \sim U(0,
        0.05)$, and $\descendantThetaMean{} \sim U(0, 0.05)$.
    \item The \modelUshaped model with the U-shaped prior of \msb on unordered
        divergence models, but with exponential priors on divergence-time and
        demographic parameters; $\divTimeIndexVector \sim \priorOld$,
        $\divTime{} \sim Exp(mean=2.887)$, $\ancestralTheta{} \sim
        Exp(mean=0.025)$, $\descendantTheta{1}{} \sim Exp(mean=0.025)$, and
        $\descendantTheta{2}{} \sim Exp(mean=0.025)$.
    \item The \modelUniform model with a uniform prior over unordered
        divergence models and exponential priors on divergence-time and
        demographic parameters; $\divTimeIndexVector \sim \priorUniform$,
        $\divTime{} \sim Exp(mean=2.887)$, $\ancestralTheta{} \sim
        Exp(mean=0.025)$, $\descendantTheta{1}{} \sim Exp(mean=0.025)$, and
        $\descendantTheta{2}{} \sim Exp(mean=0.025)$.
    \item The \modelDPP model with a Dirichlet-process prior on ordered
        divergence models and exponential priors on divergence-time and
        demographic parameters; $\divTimeIndexVector \sim \priorDPP{\sim
        Gamma(2,2)}$, $\divTime{} \sim Exp(mean=2.887)$, $\ancestralTheta{}
        \sim Exp(mean=0.025)$, $\descendantTheta{1}{} \sim Exp(mean=0.025)$,
        and $\descendantTheta{2}{} \sim Exp(mean=0.025)$.
\end{enumerate}
I selected the exponential prior on divergence time used in models \modelDPP,
\modelUniform, and \modelUshaped to have the same variance as the uniform prior
in model \modelOld.
I selected the exponential prior on population size used in models \modelDPP,
\modelUniform, and \modelUniform to have the same mean as the uniform prior in
model \modelOld, so that all four models have the same \myTheta{C} and thus the
same units of time.
All of the models were the same in other respects, with three free \myTheta{}
parameters for each population pair, two uniformly distributed ($beta(1,1)$)
\bottleScalar{}{} parameters per pair, no migration, no recombination, and
resorting of taxon-specific summary statistics by $\pi_b$ (i.e., sampling
unordered divergence models).
For all simulations, I used a data structure of eight population pairs, with a
single 1000 base-pair locus sampled from 10 individuals from each population.

@@

For each set of 50,000 simulated datasets, I used the posterior estimates
to assess the model-choice behavior of each model.
I did this by assigning the 50,000 estimates of the posterior probability
of one-divergence event to 20 bins of width 0.05, and plotted
the estimated posterior probability of each bin against the proportion of
replicates in that bin with a true value consistent with one divergence
event \cite{Huelsenbeck2004,Oaks2012}.
If the method is unbiased, the points should fall near the identity line,
whereas any trend away from the identity line indicates bias.
I did this using two criteria for the one-divergence model:
(1) the number of divergence-time parameters ($\divTimeNum = 1$) and
(2) the dispersion index of divergence times ($\divTimeDispersion < 0.01$).
For the latter, $\divTimeDispersion < 0.01$ has been commonly used as an
arbitrary criterion for a single ``simultaneous'' divergence event
(e.g., \cite{Hickerson2006,Leache2007,Stone2012}).
I focused on the one-divergence model to assess model-choice behavior, because
it is often of biogeographic interest and is easily comparable among
the three different priors used on divergence models.

@@

\subsection{Assessing power}
I evaluated the power of the same four models (Table~\ref{tabPriors}) to
detect random variation in divergence times using methods similar to
Oaks et al.\ \cite{Oaks2012}.
For all power simulations, I used a data structure identical to that of the
empirical dataset of Philippine vertebrates analyzed by Oaks et al.\
\cite{Oaks2012}, which consists of 22 pairs of populations.
Due to the larger number of pairs, I used a different hyperprior on the
concentration parameter for the \modelDPP model; I used a prior of
$\divTimeIndexVector \sim \priorDPP{\sim Gamma(1.5,18.1)}$ over divergence
models for the model \modelDPP.
All other aspects of the four models in Table~\ref{tabPriors} were identical to
those used in the validation analyses described above.
For each of the four models, I generated $2\e6$ samples from the prior.

@@

\subsection{Validation analyses: Model-choice accuracy}
The \msb model, and my modification of it, is a model-choice method
with the primary purpose of estimating the probabilities of models
of divergence across taxa.
Thus, it is critical to assess the method's ability to accurately
estimate the posterior probabilities of divergence models.
Consistent with the findings of Oaks et al.\ \cite{Oaks2012}, my results
demonstrate that the unadjusted estimates of divergence-model posterior
probabilities are generally more accurate than regression-adjusted estimates
(compare the plots along the upper-left to lower-right diagonal for
Figure~\ref{figValidationModelChoicePsi} versus
S\ref{figValidationModelChoicePsiGlm} and
Figure~\ref{figValidationModelChoiceOmega} versus
S\ref{figValidationModelChoiceOmegaGlm}).
Regression adjustment results in biased estimates of the posterior probability
of the one-divergence model when all model assumptions are satisfied, which is
well illustrated in Figure S\ref{figValidationModelChoiceOmegaGlm}.
As a result, I will focus my discussion of the results on the unadjusted
estimates.
